Bridging the foundational gap: Updating algebraic geometry in face of current challenges regarding formalizability, constructivity and predicativity

### 2021-11-23 slides-dagstuhl2021.tex
Finished slides

\title{Bridging the foundational gap: updating algebraic geometry in
face of current challenges regarding formalizability, constructivity and
predicativity}
\author{Ingo Blechschmidt}
\date{November 23th, 2021}

@@

  \begin{tikzpicture}
    \def\R{8pt}
    \node (pretitle) {Bridging the foundational gap:};
    \node (title) [below of=pretitle] {\vbox{\vspace*{-0.5em}updating \hil{algebraic
    geometry} in face of \hil{current challenges} regarding
    \hil{formalizability}, \hil{constructivity} and \hil{predicativity}}};
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple]
        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}

@@

  \subhead{Examples}
  \begin{enumerate}
    \item $\Spec(k[X_1,\ldots,X_n]) = \AA^n$.
    \item $\Spec(k[X,Y]/(Y-X^2)) = \text{standard parabola}$.
    \item Gluing
    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
      \draw (0,0) circle (1);
      \draw [fill=white] (90:1) circle (0.2);
    \end{tikzpicture}
    with
    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
      \draw (0,0) circle (1);
      \draw [fill=white] (270:1) circle (0.2);
    \end{tikzpicture}
    along
    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
      \draw (0,0) circle (1);
      \draw [fill=white] (90:1) circle (0.2);
      \draw [fill=white] (270:1) circle (0.2);
    \end{tikzpicture}
    yields
    $\PP^1$.
  \end{enumerate}
  \bigskip

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.95cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
\begin{frame}{Transfinite methods}
  The current presentation of algebraic geometry hinges on:
  \begin{itemize}
    \item large structures: classes, large categories, universes, \ldots
    \item powersets
    \item law of excluded middle
    \item axiom of choice
  \end{itemize}
  \bigskip

  \subhead{despite:}
  \begin{enumerate}
    \item subject matter (in part) very concrete
    \item practical computer algebra systems for computations
    \item well-developed constructive algebra
  \end{enumerate}
\end{frame}}

@@

  \subhead{Schemes and scheme morphisms}
  \only<1-2>{
  \begin{itemize}
    \setlength\itemsep{-0.2em}
    %\item locally ringed spaces (or locales, distributive lattices, sites,
    %toposes, arithmetic universes), formal geometries
    \item locally ringed \ldots \\
    \ldots{} spaces \tabto{4cm}-- \textit{unconstructive, hard to formalize} \\
    \ldots{} locales \tabto{4cm}-- \textit{impredicative, superfluous opens} \\
    \ldots{} distributive lattices \tabto{4cm}-- \textit{need extension from basis} \\
    \ldots{} sites \tabto{4cm}-- \textit{current favorite! morphisms intricate} \\
    \ldots{} toposes \tabto{4cm}-- \textit{large structure} \\
    \ldots{} arithmetic universes \tabto{4cm}\mbox{-- \textit{a bit better;
    issue with relative spectrum}}
    \item formal geometries \tabto{4cm}-- \textit{can be regarded as sites}
    \item functor of points \tabto{4cm}-- \textit{large structure, issue with
    sizes or \\ \qquad
    with schemes not of finite presentation over the base ring}
    \item formal gluing data \tabto{4cm}-- \textit{morphisms intricate}
  \end{itemize}
  \textbf{A surprise of uncertain import:}
  \pause
  Internally to the big Zariski topos of a base scheme, the Zariski spectrum of
  a finitely presented algebra does have enough points! [Cherubini--Coquand]}
  \bigskip
  \pause

  \subhead{Zariski cohomology}
  \begin{itemize}
    \setlength\itemsep{-0.2em}
    \small
    \item Čech cohomology \tabto{4.5cm}-- \textit{ad hoc, but fine for}
    \tabto{4.5cm}\phantom{--} \textit{quasicompact separated schemes}
    \item injective resolutions \tabto{4.5cm}-- \textit{hopelessly unconstructive}
    \item flabby resolutions \tabto{4.5cm}-- \textit{probably unconstructive}
    \item pointwise Kan extensions \tabto{4.5cm}-- \textit{fine!}
  \end{itemize}
  \bigskip

@@

\section{The generic model}

\subsection{The generic ring}

\begin{frame}{The generic ring}
  \visible<2->{``}Let~$R$ be a ring.\visible<2->{''} \visible<2->{-- Which ring does this phrase refer to?}

  \vspace*{-0.6em}
  \begin{center}
    \begin{tikzpicture}[ultra thick, node distance=7mm]
      \node[rectangle, rounded corners=1pt, inner sep=5pt, draw=lime!80, fill=lime!40] (a) {$\ZZ$};
      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of a] (b) {$\FF_2$};
      \node[rectangle, rounded corners=1pt, draw=lime!80, fill=lime!40, right=of b] (c) {$\QQ[X]$};
      \node[regular polygon, regular polygon sides=6, draw=orange!80, fill=orange!40, right=of c, rounded corners=1pt] (d) {$\RR$};
      \node[regular polygon, regular polygon sides=5, draw=pink!80, fill=pink!40, right=of d, rounded corners=1pt, inner sep=0cm] (e) {$\O_X$};
      \visible<3->{\node[star, rounded corners=1pt, star points=10, inner
      sep=2pt, draw=purple!80, fill=purple!30, right=of e] {$\AA$};}
    \end{tikzpicture}
  \end{center}
  \vspace*{-0.6em}

  \pause
  \pause
  \pause

  \justifying
  \textbf{Thm.} For any$^\star$ property~$P$ of rings, the following are
  equivalent: \\[0em]
  \begin{enumerate}
    \item The \hil{generic ring}~$\AA$ has property~$P$.
    \item Every$^\star$ ring has property~$P$.
    \item The ring axioms entail property~$P$.
  \end{enumerate}
  \pause

  \textbf{Example A.} For any~$x,y,z \in \AA$, $x + (y + z) = (x + y) + z$.
  \pause

  \textbf{Example B.} \only<6>{Is~$1 + 1 = 0$ in~$\AA$?}%
  \pause
  \only<7->{It is \hil{not the case} that~$1 + 1 = 0$ in~$\AA$. \only<8->{But also:}} \\
  \pause
  \phantom{\textbf{Example B.}} It is \hil{not the case} that~$1 + 1 \neq 0$ in~$\AA$.
  \pause

  \textbf{Example C (Anders Kock).} The generic ring is a \hil{field}:
  \[ \forall x \in \AA\_ \bigl((x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y
  \in \AA\_ xy = 1)\bigr). \]

  \pause
  \vspace*{-0.6em}
  \hil{Hence:} When verifying a coherent sequent for all rings, can without loss of
  generality assume the field condition.
\end{frame}

\begin{frame}{A selection of noncoherent sequents}
  The \hil{generic object}~$\MM$ validates:
  \begin{enumerate}
    \item $\forall x,y \in \MM\_ \neg\neg(x = y)$.
    \item $\forall x_1,\ldots,x_n \in \MM\_ \neg \forall y \in \MM\_
    y = x_1 \vee \cdots \vee y = x_n$.
  \end{enumerate}

  The \hil{generic ring}~$\AA$ validates:
  \begin{enumerate}
    \item $\forall x \in \AA\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y \in \AA\_ xy = 1)$.
    \item $\forall x \in \AA\_ \neg\neg(x = 0)$.
  \end{enumerate}

  The \hil{generic local ring}~$\AA'$ validates:
  \begin{enumerate}
    \item $\forall x \in \AA'\_ (x = 0 \Rightarrow 1 = 0) \Rightarrow (\exists y \in \AA'\_ xy = 1)$.
    \item $\neg \forall x \in \AA'\_ \neg\neg(x = 0)$.
    \item $\forall f \in \AA'[X]_{\text{degree} > 0}\_ \neg\neg \exists x \in \AA'\_ f(x) = 0$.
  \end{enumerate}
\end{frame}

\begin{frame}{An application in commutative algebra}
  \justifying
  Let~$A$ be a reduced ring ($x^n = 0 \Rightarrow x = 0$).
  Let~$\pp$ be the \hil{generic prime ideal$^\star$} of~$A$.
  Then~$A_\pp \defeq A[\pp^{-1}]$ validates:

  {\vspace*{-1.2em}
  \setbeamercolor{block body}{bg=red!30}
  \setbeamercolor{structure}{fg=purple}
  \begin{varblock}{\textwidth}{}
    $A_\pp$ is a \hil{field}: $\forall x \in A_\pp\_ (\neg(\exists y \in A_\pp\_
    xy = 1) \Rightarrow x = 0)$.

    $A_\pp$ has \hil{$\boldsymbol{\neg\neg}$-stable equality}:
    $\forall x,y \in A_\pp\_ \neg\neg(x = y) \Rightarrow x = y$.

    \mbox{$A_\pp$ is \hil{anonymously Noetherian}.}\\[-1.2em]
  \end{varblock}}

  This observation unlocks a short and conceptual proof of Grothen\-dieck's
  \hil{generic freeness lemma} in algebraic geometry.

  \mbox{\textbf{Thm.} (baby freeness) Let~$M$ be an~$A$-module. Then~\bignumber{1}
  implies~\bignumber{3}.}

  {\small
  \begin{tabular}{@{}lc@{\ }l@{}}
    \bignumber{1} $M$ is finitely generated &($\Longleftrightarrow$&
    $M_\pp$ is finitely generated) \\
    \bignumber{2} $M$ is locally free &($\Longleftrightarrow$&
    $M_\pp$ is free) \\
    \bignumber{3} $M$ is locally free \emph{on a dense open} &($\Longleftrightarrow$&
    $M_\pp$ is \emph{not not} free) \\
  \end{tabular}\par}

  \textbf{Proof.} Elementary linear algebra over~$A_\pp$. \qed
\end{frame}

\begin{frame}{A systematic source}
  \begin{center}
    \includegraphics[width=0.8\textwidth]{wraith-some-recent-developments-in-topos-theory} \\
    \scriptsize
    Gavin Wraith. \emph{Some recent developments in topos theory.} \\ In:
    Proc.\@ of the ICM (Helsinki, 1978).
  \end{center}

  \textbf{Thm. (Nullstellensatz):} The generic~$\TT$-model~$U_T$ validates:
  For any coherent sequent~$\sigma$,
  \[
    \text{$\sigma$ holds for $U_\TT$} \quad\textcolor{purple}{\Longleftrightarrow}\quad
    \text{$\ul{\TT}/U_\TT$ proves $\sigma$.}
  \]

  \justifying
  \textbf{Thm. (universality):} The generic~$\TT$-model validates a first-order
  formula~$P$ if and only if~$P$ is intuitionistically deducible from the
  axioms of~$\TT$ and the Nullstellensatz.
\end{frame}

\begin{frame}{Arithmetic universes}
  Places where we can do mathematics (among others):
  \begin{multicols}{2}
    \begin{enumerate}
      \item $\Set$ (sets)
      \item $\Eff$ (data types) \columnbreak
      \item \mbox{$\mathrm{sSet}$ (simplicial sets)}
      \item \mbox{$\Sh(X)$ (sheaves over~$X$)}
    \end{enumerate}
  \end{multicols}
  \pause
  \vspace*{-0.6em}
  These are examples for \hil{arithmetic universes}.
  \pause

  \justifying
  \textbf{Definition.} An \emph{arithmetic universe} is a category with finite
  limits (``$\times$''), stable finite disjoint coproducts (``$\amalg$''),
  stable ef{}fective quotients (``$X/{\sim}$'') and parametrized list objects
  (``$\NN$'', ``$\mathrm{List}(X)$'').

  \textbf{Thm.} Any statement which is provable in \hil{predicative constructive
  mathematics} (no powersets, no $\varphi \vee \neg\varphi$, no
  \mbox{$\neg\neg\varphi \Rightarrow \varphi$}, no axiom of choice) is true in any
  arithmetic universe.
  \pause

  Further examples:

  \begin{enumerate}
    \addtocounter{enumi}{4}
    \item the \hil{initial} arithmetic universe
    \item the \hil{classifying} arithmetic universe for the theory of rings
  \end{enumerate}
\end{frame}

>>>>

\title{Bridging the foundational gap: Updating algebraic geometry in
face of current challenges regarding formalizability, constructivity and
predicativity}
\author{Ingo Blechschmidt}
\date{November 23th, 2021}

@@

  \begin{tikzpicture}
    \def\R{8pt}
    \node (pretitle) {Bridging the foundational gap:};
    \node (title) [below of=pretitle] {\vbox{\vspace*{-0.5em}Updating \hil{algebraic
    geometry} in face of \hil{current challenges} regarding
    \hil{formalizability}, \hil{constructivity} and \hil{predicativity}}};
    \begin{pgfonlayer}{background}
      \draw[decorate, very thick, draw=mypurple]
        ($(title.south west) + (\R, 0)$) arc(270:180:\R) --
        ($(title.north west) + (0, -\R)$) arc(180:90:\R) --
        ($(title.north east) + (-\R, 0)$) arc(90:0:\R) --
        ($(title.south east) + (0, \R)$) arc(0:-90:\R) --
        cycle;
    \end{pgfonlayer}
  \end{tikzpicture}

@@

  \subhead{Examples}
  \begin{enumerate}
    \item $\Spec(k[X_1,\ldots,X_n]) = \AA^n$.
    \item $\Spec(k[X,Y]/(Y-X^2)) = \text{standard parabola}$.
    \item Gluing
    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
      \draw (0,0) circle (1);
      \draw [fill=white] (90:1) circle (0.3);
    \end{tikzpicture}
    with
    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
      \draw (0,0) circle (1);
      \draw [fill=white] (270:1) circle (0.3);
    \end{tikzpicture}
    along
    \begin{tikzpicture}[scale=0.2,baseline={([yshift=-.6ex]current bounding box.center)}]
      \draw (0,0) circle (1);
      \draw [fill=white] (90:1) circle (0.3);
      \draw [fill=white] (270:1) circle (0.3);
    \end{tikzpicture}
    yields
    $\PP^1$.
  \end{enumerate}
  \bigskip

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{3.95cm}\includegraphics[width=\paperwidth]{staircase}\end{minipage}}
\begin{frame}{Transfinite methods}
  The standard presentation of algebraic geometry hinges on:
  \begin{itemize}
    \item large structures: classes, large categories, universes, \ldots
    \item powersets
    \item law of excluded middle
    \item axiom of choice
  \end{itemize}
  \bigskip

  \subhead{despite:}
  \begin{enumerate}
    \item subject matter (in part) very concrete
    \item computer algebra systems for computations practical
    \item constructive algebra well-established
    \item high-level proofs often constructive
  \end{enumerate}
\end{frame}}

@@

  \subhead{Schemes and scheme morphisms}
  \only<1-2>{
  \begin{itemize}
    \setlength\itemsep{-0.2em}
    %\item locally ringed spaces (or locales, distributive lattices, sites,
    %toposes, arithmetic universes), formal geometries
    \item locally ringed \ldots \\
    \ldots{} topological spaces \tabto{4cm}-- \textit{unconstructive, hard to formalize} \\
    \ldots{} locales \tabto{4cm}-- \textit{impredicative, superfluous opens} \\
    \ldots{} distributive lattices \tabto{4cm}-- \textit{need extension from basis} \\
    \ldots{} sites \tabto{4cm}-- \textit{current favorite! morphisms intricate} \\
    \ldots{} toposes \tabto{4cm}-- \textit{large structure} \\
    \ldots{} arithmetic universes \tabto{4cm}\mbox{-- \textit{a bit better;
    issue with relative spectrum}}
    \item formal geometries \tabto{4cm}-- \textit{can be regarded as sites}
    \item functor of points \tabto{4cm}-- \textit{large structure, issue with
    sizes or \\ \qquad
    with schemes not of finite presentation over the base ring}
    \item formal gluing data \tabto{4cm}-- \textit{morphisms intricate}
  \end{itemize}
  \textbf{A surprise of uncertain import:}
  \pause
  Internally to the big Zariski topos of a base scheme, the Zariski spectrum of
  a finitely presented algebra does have enough points! [Cherubini--Coquand]}
  \bigskip
  \pause

  \subhead{Zariski cohomology}
  \begin{itemize}
    \setlength\itemsep{-0.2em}
    \small
    \item Čech cohomology \tabto{4.5cm}-- \textit{ad hoc, but fine for}
    \tabto{4.5cm}\phantom{--} \textit{quasicompact separated schemes}
    \item injective resolutions \tabto{4.5cm}-- \textit{hopelessly unconstructive}
    \item dynamical injectives \tabto{4.5cm}-- \textit{??}
    \item flabby resolutions \tabto{4.5cm}-- \textit{probably unconstructive}
    \item pointwise Kan extensions \tabto{4.5cm}-- \textit{fine! partially
    defined $\mathbb{R}\Gamma$;} \tabto{4.5cm}-- \textit{existence
    verified for} \tabto{4.5cm}\phantom{--} \textit{quasicompact separated schemes;}
    \tabto{4.5cm}\phantom{--} \textit{hyper coverings?}
  \end{itemize}

@@
